Communication complexity: Chapter 5 \\ Information complexity

### 2019-02-06 CS860-CommunicationComplexity/assignments/Chapter5.tex
Fixed p1

Let $\Pi$ be a distribution on protocols such that $\cost(\Pi) = R_\epsilon(f)$. That is,
$$R_\epsilon(f) = \max_{(x,y) \in \calX\times\calY}\max_{\pi \sim \Pi}\height(\pi(x,y))$$ 
where $\height(\pi(x,y))$ denotes the length of the path from the root to leaf of the protocol tree $\pi$ dictated by inputs $(x,y)$. We use $\height(\pi)$ to denote $\max_{(x,y) \in \calX\times\calY}\height(\pi(x,y))$. Let $\pi^*$ be the protocol tree achieving cost $\R_\epsilon(f)$. Thus we have
$$R_\epsilon(f) = \height(\pi^*) \geq \E_{(x,y) \sim \mu} [\height(\pi^*(x,y))].$$
Let $\Gamma^{\pi^*}_{X,Y}(r)$ denote the shared random string part of $\Gamma^{\pi^*}_{X,Y}$, and let $\Gamma^{\pi^*}_{X,Y}(c)$ denote the communication between Alice and Bob of the transcript. We claim that
$$\E_{(x,y) \sim \mu} [\height(\pi^*(x,y))] \geq H(\Gamma^{\pi^*}_{X,Y}(c)).$$
To see this claim is sufficient, observe that it implies 
\begin{align*}
R_\epsilon(f) &\geq H(\Gamma^{\pi^*}_{X,Y}(c)) \\
&\geq I(XY; \Gamma^{\pi^*}_{X,Y}(c) \mid \Gamma^{\pi^*}_{X,Y}(r)) \\
&= I(XY;\Gamma^{\pi^*}_{X,Y}(c) \mid \Gamma^{\pi^*}_{X,Y}(r)) + I(XY;\Gamma^{\pi^*}_{X,Y}(r))\\
&= I(\Gamma^{\pi^*}_{X,Y}; XY) \\
&= \icostext_\mu(\Gamma^{\pi^*}_{X,Y})\\
&\geq \ICext_{\mu,\epsilon}(f).
\end{align*}

Now we proceed to prove the claim. Let $L$ be the set of leaves of the protocol tree $\pi^*$. We define a distribution $\mu'$ on $L$ as follows: for each $\ell \in L$, we let
 $$\mu'(\ell = \sum_{(x,y) \in \calX\times\calY : (x,y)\text{ induces a path to } \ell} \mu(x,y).$$ 
 Let $h_\ell = \height(x,y)$ where $(x,y)$ induces a path to $\ell$. i.e. $h_\ell$ is the lenght of the path from the root of $\pi^*$ to $\ell$ (this path is taken on input $(x,y)$). We have
\begin{align*}
\E_{(x,y)\sim \mu}[h_{x,y}] &=\sum_{\ell \in L} \mu'(\ell) h_\ell \\
&=\sum_{\ell \in L} \mu'(\ell)\left(\log(\frac{1}{\mu'(\ell)})-\log(\frac{1}{\mu'(\ell)}) +\log(2^{h_\ell})\right)\\
&=\sum_{\ell\in L} \mu'(\ell)\left(\log(\frac{1}{\mu'(\ell)})-\log(\frac{2^{-h_\ell}}{\mu'(\ell)}))\right) \\
&=H(\Gamma^{\pi^*}_{X,Y}(c)) - \sum_{\ell\in L}\mu'(\ell)\log(\frac{2^{-h_\ell}}{\mu'(\ell)}) \\
&\geq H(\Gamma^{\pi^*}_{X,Y}(c)) - \log(\sum_{\ell\in L} 2^{-h_\ell})
\end{align*}
where the last line follows by applying the convexity of $\log$. The sum inside the logarithm, $\sum_{\ell \in L} 2^{-h_\ell}$ is at most $1$, completing the proof. This is a basic fact about binary trees. We'll give a brief proof for completeness.
Proceed by induction on $|L|$. If $|L| = 1$, the result is immediate. For induction suppose the result holds for binary trees smaller that $\pi^*$. Let $r$ be the root of $\pi^*$. Split $\pi^*$ into two subtrees $T_1$ and $T_2$ by deleting $r$. Now observe that
$$\sum_{\ell \in L} 2^{-h_\ell} = \frac{1}{2}\sum_{\ell \in L \cap T_1} + \frac{1}{2}\sum_{\ell \in L \cap T_2} \leq \frac{1}{2}+\frac{1}{2} = 1.$$
With the first equality following since joining $T_1$ and $T_2$ by root $r$, adds $1$ to length of the path to each leaf, and the inequality follows by induction.
\end{proof}

>>>>

Let $\Pi$ be a distribution on protocols such that $\cost(\Pi) = R_\epsilon(f)$. That is,
$$R_\epsilon(f) = \max_{(x,y) \in \calX\times\calY}\max_{\pi \sim \Pi}\height(\pi(x,y))$$ 
where $\height(\pi(x,y))$ denotes the length of the path from the root to leaf of the protocol tree $\pi$ dictated by inputs $(x,y)$. We use $\height(\pi)$ to denote $\max_{(x,y) \in \calX\times\calY}\height(\pi(x,y))$. Let $\pi^*$ be the protocol tree achieving cost $\R_\epsilon(f)$. Thus we have
$$R_\epsilon(f) = \height(\pi^*) \geq \E_{(x,y) \sim \mu} [\height(\pi^*(x,y))].$$
Let $\Gamma^{\pi^*}_{X,Y}(r)$ denote the shared random string part of $\Gamma^{\pi^*}_{X,Y}$, and let $\Gamma^{\pi^*}_{X,Y}(c)$ denote the communication between Alice and Bob of the transcript. We claim that
$$\E_{(x,y) \sim \mu} [\height(\pi^*(x,y))] \geq H(\Gamma^{\pi^*}_{X,Y}(c)).$$
To see this claim is sufficient, observe that it implies 
\begin{align*}
R_\epsilon(f) &\geq H(\Gamma^{\pi^*}_{X,Y}(c)) \\
&\geq I(XY; \Gamma^{\pi^*}_{X,Y}(c) \mid \Gamma^{\pi^*}_{X,Y}(r)) \\
&= I(XY ;\Gamma^{\pi^*}_{X,Y}) \\
&= \icostext_\mu(\Gamma^{\pi^*}_{X,Y})\\
&\geq \ICext_{\mu,\epsilon}(f).
\end{align*}

Now we proceed to prove the claim. Let $L$ be the set of leaves of the protocol tree $\pi^*$. We define a distribution $\mu'$ on $L$ as follows: for each $\ell \in L$, we let
 $$\mu'(\ell) = \sum_{(x,y) \in \calX\times\calY : (x,y)\text{ induces a path to } \ell} \mu(x,y).$$ 
 Let $h_\ell = \height(x,y)$ where $(x,y)$ induces a path to $\ell$. i.e. $h_\ell$ is the length of the path from the root of $\pi^*$ to $\ell$ (this path is taken on input $(x,y)$). We have
\begin{align*}
\E_{(x,y)\sim \mu}[h_{x,y}] &=\sum_{\ell \in L} \mu'(\ell) h_\ell \\
&=\sum_{\ell \in L} \mu'(\ell)\left(\log(\frac{1}{\mu'(\ell)})-\log(\frac{1}{\mu'(\ell)}) +\log(2^{h_\ell})\right)\\
&=\sum_{\ell\in L} \mu'(\ell)\left(\log(\frac{1}{\mu'(\ell)})-\log(\frac{2^{-h_\ell}}{\mu'(\ell)}))\right) \\
&=H(\Gamma^{\pi^*}_{X,Y}(c)) - \sum_{\ell\in L}\mu'(\ell)\log(\frac{2^{-h_\ell}}{\mu'(\ell)}) \\
&\geq H(\Gamma^{\pi^*}_{X,Y}(c)) - \log(\sum_{\ell\in L} 2^{-h_\ell})
\end{align*}
where the last line follows by applying the convexity of $\log$. The sum inside the logarithm, $\sum_{\ell \in L} 2^{-h_\ell}$ is at most $1$, completing the proof. This is a basic fact about binary trees. We'll give a brief proof for completeness.
Proceed by induction on $|L|$. If $|L| = 1$, the result is immediate. For induction suppose the result holds for binary trees smaller that $\pi^*$. Let $r$ be the root of $\pi^*$. Split $\pi^*$ into two subtrees $T_1$ and $T_2$ by deleting $r$. Now observe that
$$\sum_{\ell \in L} 2^{-h_\ell} = \frac{1}{2}\sum_{\ell \in L \cap T_1} + \frac{1}{2}\sum_{\ell \in L \cap T_2} \leq \frac{1}{2}+\frac{1}{2} = 1.$$
With the first equality following since joining $T_1$ and $T_2$ by root $r$, adds $1$ to length of the path to each leaf, and the inequality follows by induction.
\end{proof}
